Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-01 Manuscript/02-Method.Rmd
Fix the table and figure placement
Revise for ENAR submission

Following the GLM notation introduced in Section \ref{sec:intro}, we have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
\end{equation}
and the data distribution is
\begin{equation}
f(\bs Y = \bs y| \bs \beta, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\bs \beta, \phi),\nonumber
\end{equation}
with smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix, i.e. the design matrix derived from the smoothing function $B_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its corresponding smoothing penalty matrix is denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable such that $B_j(x_j) = \bs \beta_j^T \bs X_j$. With slight abuse of notation, we denote vectors and matrices in bold fonts $\bs \beta, \bs X$ with conformable dimensions, where scalar and random variables are denoted in with unbold fonts $\beta, X$. The matrix transposing operation is denoted with a superscript $^T$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models. The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function. However, the direct integration of smoothing penalty with sparsity penalty is not obvious. Marra and  Wood [@Marra2011] proposed a re-parameterization procedure to accommodate the smoothing penalty implicitly. Given the smoothing penalty matrix $\bs S_j$ is symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where the matrix $\bs D$ is diagonal with the eigenvalues arranged in the ascending order. To note, $\bs D$ can contain elements of zeros on the diagonal, where the zeros are associated with the linear space of the smoothing function. For the most popular smoothing function, cubic splines, the dimension of the linear space is one. Hereafter, we focus on discussing a uni-dimensional linear space for simplicity; however, it generalizes easily to the cases where the linear space is multidimensional. We further write the orthonormal matrix $\bs U \equiv \begin{bmatrix} \bs U^0 : \bs U^{\tp}\end{bmatrix}$ containing the eigenvectors as columns in the corresponding order to $\bs D$. That is, $\bs U$ contains the eigenvectors $U^0$ with zero eigenvalues for the linear space and $\bs U^{\tp}$ contains the eigenvectors (as columns) for the non-zero eigenvalues, i.e. the non-linear space. We multiply the basis function matrix $\bs X$ with the orthonormal matrix $\bs U$ for the new design matrix ${\bs X}^\repa = \bs X \bs U \equiv \begin{bmatrix} X^0 : \bs X^{\tp} \end{bmatrix}$. An additional scaling step is imposed on $\bs X^{\tp}$ by the non-zero eigenvalues of $\bs D$ such that the new basis function matrix $\bs X^\ast$ can receive uniform penalty on each of its dimensions. With slight abuse of the notation, we drop the superscript $^\repa$ and denote $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable after the re-parameterization. A spline function can be expressed in the matrix form
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\bs \beta_j^\tp}^T \bs X_j^\tp)\right],
\end{equation}
where the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space.

The re-parameterization step sets up the foundation of the proposed model and provides three-fold benefits. First of all, the re-parameterization integrates the smoothing penalty into the design matrix, and encourages models to properly smooth the nonlinear function in addition to the sparse penalty for function selection. Secondly, the eigendecomposition of the smoothing penalty allows the isolation of the linear from the nonlinear space, improving the feasibility of bi-level function selection and inference. The eigendecomposition facilitates the construction of orthonormal design matrix, which makes imposing independent priors on the coefficients possible. This reduces the computational complexity compared to using a multivariate priors, and greatly broadens the choices of priors and further model choices. Last but not least, the scaling step of $\bs X^\tp$ further simplifies the choice of priors, from column-specific priors to a unified prior.

## Spike-and-Slab Smooth Priors

The family of spike-and-slab (SS) priors regression models dominates Bayesian high-dimensional analysis for its utility in outcome prediction and variable selection. We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically
$$
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
$$
The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see George and McCulloch [@George1993; @George1997]; Chipman [@Chipman1996] for grouped variables; Brown et al. [@Brown1998] for multivariate outcomes; Ishwaran and Rao [@Ishwaran2005]; Clyde and George [@Clyde2004] and reference therein.

<!-- Comments -->
<!-- Even the conjugate normal prior of $\bs \beta$ provides analytic simplicity, there is no closed form solution for the posterior density of $\bs \gamma$. @George1993 described a Gibbs sampling algorithm to approximate the posterior distribution of $\bs \gamma$ for variable selection. The same authors later compared SSVS with other Bayesian variable selection approaches, providing some empirical suggestions on the advantages and disadvantages of various formulations and estimation algorithms. [@George1997] Among the various problem SSVS addresses, high-dimensionality is not one of them, due to the intensive computation needs of MCMC algorithms. @George1997 suggested MCMC algorithms works well for medium size of predictors (p=25), which is not feasible for high-dimensional data, where the number of predictors easily exceeds 100. -->

The major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018]. Particularly, the development of spike-and-slab Lasso  substantially improves the scalability of SS models, setting up the theoretical foundation for generalized models in -omics data analysis [@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The spike-and-slab Lasso (SSL) prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation}
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$. The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection.

### Spike-and-Slab Lasso Prior

We introduce a novel prior for GAMs, particularly for high-dimensional nonlinear modeling with bi-level selection. The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} X^0_j : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j = \begin{bmatrix} \beta_j : \bs \beta_j^\tp \end{bmatrix}$. Specifically, we impose independent group priors on the linear space coefficients and on the nonlinear space coefficients respectively, 
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j
\end{align}
where $\gamma_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}$ are two latent indicator variables, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}. To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], as the $\beta_j$ and $\bs \beta^\tp_j$ have different indicator variables $\gamma_j$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta_{jk} \in \bs \beta^\tp_j$ has its own latent indicator $\gamma_{jk}$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior.

The re-parameterization introduced in Section \ref{sec:BHAM} grants the validity of the proposed prior. First of all, the smoothing function bases are linear dependent and ask for extra attention. The eigeondecomposition remedies the problem and hence our prior can be set to be conditionally independent. Secondly, the eigenvalue scaling provides a panacea to allow unified scale parameters for all bases of all smoothing functions of variables.

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors on $\gamma_j, \gamma^\tp_j$ to allow local adaption of the shrinkage using a Bernoulli distribution, written as binomial distribution of one trial. The two indicators of the $j$th predictor, $\gamma_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^\tp | \theta_j \sim Bin(1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1), i.e. $a = 1, b = 1$. When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM-SSL, and visually presented in Figure \ref{fig:SSprior}.

\begin{figure}
\centering
\begin{tikzpicture} [
staticCompo/.style = {rectangle, minimum width=1cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
outCome/.style={ellipse, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
mymatrix/.style={matrix of nodes, nodes=outCome, row sep=1em},
PriorBoarder/.style={rectangle, minimum width=5cm, minimum height=10cm, text centered, fill=lightgray!30},
background/.style={rectangle, fill=gray!10,inner sep=0.2cm, rounded corners=5mm}
]

@@

\end{tikzpicture}

\caption{Directed acyclic grpah of BHAM-SSL model with parameter expansion. Elliposes are stochastic nodes, rectangles and are deterministic nodes. }
\label{fig:SSprior}
\end{figure}

### Other Priors

With the re-parameterization step of the basis function matrix $\bs X$, it is possible to generalized the SSL prior to other priors, for example normal priors for ridge-type regularization and mixture normal prior for spike-and-slab regularization. These priors would work better in low and medium dimensional settings where the sparse assumption is not necessary. Here we elaborate the mixture normal prior as a demonstration of applying continuous spike-and-slab prior in BHAM.

A spike-and-slab mixture normal spline prior can be expressed as 
\begin{align*}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim N(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1)\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid N(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j.
\end{align*}
Similar to the BHAM-SSL prior in Equation (\ref{eq:bham_ssl}), $0 < s_0 < s_1$ are tuning parameters and can be optimized via cross-validation. One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which causes problems when estimating the large effects. Distributions with heavier tails can be used as an alternative, for example mixture Student's $t$ distribution with small degree of freedom.

@@

The proposed models can be fitted with MCMC algorithms. Nevertheless, the computational burden of MCMC algorithms creates scalability issues. George and McCulloch[@George1997] examined the computation speed for various MCMC algorithms with spike-and-slab mixture normal priors, and suggested MCMC algorithms works well for medium size ($p$=25) of predictors with only linear effects. However, it is not feasible for high-dimensional data analysis where the number of predictors easily exceeds 100. Specific to additive models, each predictor would expand to multiple new "predictors" via basis functions, creating greater computational demands. Scheipl et al. [@Scheipl2013] demonstrated the computational demands of a MCMC algorithm for fitting spike-and-slab GAM grow exponentially as $p$ increases modestly via simulation studies. Hence, we feel compelled to develop scalable algorithms for fitting Bayesian hierarchical additive models in high-dimensional settings.

>>>>

A generalized additive model can be formulated in its matrix form as,
\begin{equation}\label{eq:gam}
E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
\end{equation}
and the data distribution is
\begin{equation}
f(\bs Y = \bs y| \bs \beta, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\bs \beta, \phi),\nonumber
\end{equation}
with the disperse parameter $\phi$ and smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix is denoted $\bs X_j$ for the variable $x_j$
<!-- . The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its  -->
with its smoothing penalty matrix denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients such that $B_j(x_j) = \bs \beta_j^T \bs X_j$. With slight abuse of notation, we denote vectors and matrices in bold fonts $\bs \beta, \bs X$ of conformable dimensions, where scalar and random variables are denoted in with unbold fonts $\beta, X$. The matrix transposing operation is denoted with a superscript $^T$. 
<!-- To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.  -->

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models. 
<!-- The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function. However, the direct integration of smoothing penalty with sparsity penalty is not obvious. -->
While it is hard to directly integrate smoothing penalty with sparsity penalty, @Marra2011 proposed a reparameterization to implement smoothing penalty in the design matrix. Given the smoothing penalty matrix $\bs S_j$ is symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where eigenvectors and eigenvalues are in the matrices $\bs U$ and $\bs D$ respectively. The zero eigenvalue and its corresponding eigenvector span the linear space of the smoothing function, which allows us to separate the linear space from the smoothing function. By multiplying the design matrix $X$ and eigenvector matrix $U$ and proper scaling with eigenvalues, we can have a new design matrix such that the smoothing function of variable $x_j$ can be written as
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
where $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable; the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space.

<!-- To note, $\bs D$ can contain elements of zeros on the diagonal, where the zeros are associated with the linear space of the smoothing function. For the most popular smoothing function, cubic splines, the dimension of the linear space is one. Hereafter, we focus on discussing a uni-dimensional linear space for simplicity; however, it generalizes easily to the cases where the linear space is multidimensional. We further write the orthonormal matrix $\bs U \equiv \begin{bmatrix} \bs U^0 : \bs U^{\tp}\end{bmatrix}$ containing the eigenvectors as columns in the corresponding order to $\bs D$. That is, $\bs U$ contains the eigenvectors $U^0$ with zero eigenvalues for the linear space and $\bs U^{\tp}$ contains the eigenvectors (as columns) for the non-zero eigenvalues, i.e. the non-linear space. We multiply the basis function matrix $\bs X$ with the orthonormal matrix $\bs U$ for the new design matrix ${\bs X}^\repa = \bs X \bs U \equiv \begin{bmatrix} X^0 : \bs X^{\tp} \end{bmatrix}$. An additional scaling step is imposed on $\bs X^{\tp}$ by the non-zero eigenvalues of $\bs D$ such that the new basis function matrix $\bs X^\ast$ can receive uniform penalty on each of its dimensions. With slight abuse of the notation, we drop the superscript $^\repa$ and denote $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable after the reparameterization. A spline function can be expressed in the matrix form -->
<!-- $$ -->
<!-- B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp , -->
<!-- $$ -->
<!-- and the generalized additive model in Equation (\ref{eq:gam}) be -->
<!-- \begin{equation}\label{eq:gam-repa} -->
<!-- E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\bs \beta_j^\tp}^T \bs X_j^\tp)\right], -->
<!-- \end{equation} -->
<!-- where the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space. -->

<!-- The reparameterization step sets up the foundation of the proposed model and provides three-fold benefits. First of all, the reparameterization integrates the smoothing penalty into the design matrix, and encourages models to properly smooth the nonlinear function in addition to the sparse penalty for function selection. Secondly, the eigendecomposition of the smoothing penalty allows the isolation of the linear from the nonlinear space, improving the feasibility of bi-level function selection and inference. The eigendecomposition facilitates the construction of orthonormal design matrix, which makes imposing independent priors on the coefficients possible. This reduces the computational complexity compared to using a multivariate priors, and greatly broadens the choices of priors and further model choices. Last but not least, the scaling step of $\bs X^\tp$ further simplifies the choice of priors, from column-specific priors to a unified prior. -->

## Spike-and-Slab Spline Priors

The family of spike-and-slab (SS) priors regression models dominates Bayesian high-dimensional analysis for its utility in outcome prediction and variable selection. 
<!-- We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically -->
<!-- $$ -->
<!--  \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta). -->
<!-- $$ -->
<!-- The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see  -->
The development includes @George1993; @George1997; @Chipman1996 for grouped variables; @Brown1998 for multivariate outcomes; @Ishwaran2005; @Clyde2004; @Rockova2014a for optimization-based model fitting algorithms and reference therein.
<!-- Comments -->
<!-- Even the conjugate normal prior of $\bs \beta$ provides analytic simplicity, there is no closed form solution for the posterior density of $\bs \gamma$. @George1993 described a Gibbs sampling algorithm to approximate the posterior distribution of $\bs \gamma$ for variable selection. The same authors later compared SSVS with other Bayesian variable selection approaches, providing some empirical suggestions on the advantages and disadvantages of various formulations and estimation algorithms. [@George1997] Among the various problem SSVS addresses, high-dimensionality is not one of them, due to the intensive computation needs of MCMC algorithms. @George1997 suggested MCMC algorithms works well for medium size of predictors (p=25), which is not feasible for high-dimensional data, where the number of predictors easily exceeds 100. -->
<!-- However, the major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018].  -->
Particularly, the development of spike-and-slab Lasso (SSL) [@Rockova2018b; @Rockova2018] substantially improves the scalability of SS models by using coordinate-wise implementations for posterior exploration. Yi and his group later developed generalized and cox SSL models for high-dimensional -omics data analysis.[@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The SSL prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{align} \label{eq:ssl}
\beta | \gamma &\sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber\\
\text{or} \qquad \beta | \gamma &\sim DE(0, S), 0 < s_0 < s_1,
\end{align}
<!-- Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density, -->
<!-- It can be aslso expressed in the form of a congregated double exponential distribution -->
<!-- \begin{equation} \label{eq:ssl} -->
<!-- \end{equation} -->
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$.

<!-- The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection. -->

<!-- ### Spike-and-Slab Lasso Prior -->
We introduce a novel SSL type prior of GAMs for high-dimensional nonlinear modeling with bi-level selection. 
<!-- The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} X^0_j : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j = \begin{bmatrix} \beta_j : \bs \beta_j^\tp \end{bmatrix}$.  -->
Specifically, we respectively impose independent group priors on the linear space coefficient $\beta_j$ and on the nonlinear space coefficients $\bs \beta^\tp_j$ of $K_j$ elements, 
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j
\end{align}
where $\gamma_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}$ are two latent indicator variables, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}.
<!-- To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], as the $\beta_j$ and $\bs \beta^\tp_j$ have different indicator variables $\gamma_j$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta_{jk} \in \bs \beta^\tp_j$ has its own latent indicator $\gamma_{jk}$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior. -->
<!-- The reparameterization introduced in Section \ref{sec:BHAM} grants the validity of the proposed prior. First of all, the smoothing function bases are linear dependent and ask for extra attention. The eigeondecomposition remedies the problem and hence our prior can be set to be conditionally independent. Secondly, the eigenvalue scaling provides a panacea to allow unified scale parameters for all bases of all smoothing functions of variables.  -->

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors on $\gamma_j, \gamma^\tp_j$ to allow local adaption of the shrinkage using a Bernoulli distribution. The two indicators of the $j$th predictor, $\gamma_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^\tp | \theta_j \sim Bin(1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1), i.e. $a = 1, b = 1$. When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM, and visually presented in Figure \ref{fig:SSprior}.

\begin{figure}
\centering
\caption{Directed acyclic grpah of BHAM-SSL model with parameter expansion. Elliposes are stochastic nodes, rectangles and are deterministic nodes. }
\label{fig:SSprior}
\begin{tikzpicture} [
staticCompo/.style = {rectangle, minimum width=1cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
outCome/.style={ellipse, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
mymatrix/.style={matrix of nodes, nodes=outCome, row sep=1em},
PriorBoarder/.style={rectangle, minimum width=5cm, minimum height=10cm, text centered, fill=lightgray!30},
background/.style={rectangle, fill=gray!10,inner sep=0.2cm, rounded corners=5mm}
]

@@

\end{tikzpicture}
\end{figure}

<!-- ### Other Priors -->

<!-- With the reparameterization step of the basis function matrix $\bs X$, it is possible to generalized the SSL prior to other priors, for example normal priors for ridge-type regularization and mixture normal prior for spike-and-slab regularization. These priors would work better in low and medium dimensional settings where the sparse assumption is not necessary. Here we elaborate the mixture normal prior as a demonstration of applying continuous spike-and-slab prior in BHAM. -->

<!-- A spike-and-slab mixture normal spline prior can be expressed as  -->
<!-- \begin{align*} -->
<!--   \beta_{j} |\gamma_{j},s_0,s_1 &\sim N(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1)\\ -->
<!--   \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid N(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j. -->
<!-- \end{align*} -->
<!-- Similar to the BHAM-SSL prior in Equation (\ref{eq:bham_ssl}), $0 < s_0 < s_1$ are tuning parameters and can be optimized via cross-validation. One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which causes problems when estimating the large effects. Distributions with heavier tails can be used as an alternative, for example mixture Student's $t$ distribution with small degree of freedom.  -->

@@

<!-- The proposed models can be fitted with MCMC algorithms. Nevertheless, the computational burden of MCMC algorithms creates scalability issues. George and McCulloch[@George1997] examined the computation speed for various MCMC algorithms with spike-and-slab mixture normal priors, and suggested MCMC algorithms works well for medium size ($p$=25) of predictors with only linear effects. However, it is not feasible for high-dimensional data analysis where the number of predictors easily exceeds 100. Specific to additive models, each predictor would expand to multiple new "predictors" via basis functions, creating greater computational demands. Scheipl et al. [@Scheipl2013] demonstrated the computational demands of a MCMC algorithm for fitting spike-and-slab GAM grow exponentially as $p$ increases modestly via simulation studies. Hence, we feel compelled to develop scalable algorithms for fitting Bayesian hierarchical additive models in high-dimensional settings.  -->
